Rotating Stars from \Kepler Observed with Gaia DR1

### 2016-11-21 paper/ms.tex
update from referee reply

\documentclass[manuscript, letterpaper]{aastex6}
%\documentclass[twocolumn]{aastex6}

@@

%\altaffiltext{1}{Corresponding author: James.Davenport@wwu.edu}
\altaffiltext{1}{Department of Physics \& Astronomy, Western Washington University, 516 High St., Bellingham, WA 98225, USA}
\altaffiltext{2}{NSF Astronomy and Astrophysics Postdoctoral Fellow}

@@

@@

\begin{figure}[]
\centering
\includegraphics[width=6in]{fig1}
\caption{
Rotation period distribution for 33,855 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots). The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$25 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}. The subsample of 1,299 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample. 
}
\label{fig:all}
\end{figure}

@@

%%%%%
\begin{figure}[]
\centering
\includegraphics[width=5.3in]{fig2}
\caption{Hertzsprung--Russell (HR) diagram using temperatures from \citet{mcquillan2014} and Gaia DR1 $G$-band absolute magnitudes for the 894 stars that pass photometric and parallax quality cuts described in the text. Points are colored by their measured \Kepler rotation periods. Two isochrones from \citet{bressan2012} are shown, with ages of 300 Myr and 4 Gyr (solid and dashed lines). Main sequence stars are selected as those between the 300 Myr isochrone and the isochrone shifted up by 1 magnitude (red line).
}
\label{fig:HR}
\end{figure}

@@

%%%%%
\begin{figure}[]
\centering
\includegraphics[width=6in]{fig3}
\caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles). 
The full \Kepler--Gaia matched sample is shown for reference (small black dots). A bimodality in rotation periods initially discovered for M dwarfs by \citet{mcquillan2013}, extends the full range of temperatures in the \Kepler--Gaia main sequence sample shown here.
A \citet{meibom2011} 600 Myr gyrochonology-isochrone (gyrochrone) traces the bimodality midpoint up to 6000 K (red solid line), but deviates from the isochrone sharply to $\sim$6200 K (red dotted line). A log-linear extrapolation of the isochrone at 6000K to 6500 K (red dashed line) continues to track the bimodality to hotter temperatures, and roughly follows a line of constant Rossby number. 
}
\label{fig:gyro}
\end{figure}

The minimum in the bimodal period distribution in Figure \ref{fig:gyro} can be traced using a gyrochonology isochrone (colloquially known as a ``gyrochrone'').  Many different studies have produced competing gyrochrone models, each with unique morphologies at the hot and cool star regimes \citep[e.g.][]{barnes2007,mamajek2008,meibom2011,angus2015}. A 600 Myr gyrochrone from \citet{meibom2011}, converted from $B-V$ color to temperature using the the transformation from \citet{sekiguchi2000}, was determined by eye to approximately trace the period minima from 3500 K to 6000 K. As shown in Figure \ref{fig:gyro}, this model (as with most gyrochronology models) turns down in period sharply for stars hotter than $\sim$6000 K. A log-linear extrapolation of the 600 Myr gyrochrone at 6000 K continues to trace the period bimodality up to 6500 K, and roughly traces a line of constant Rossby number assuming a local convective turnover timescale such as from \citet{barnes_kim2010}.

%%%%%
\begin{figure}[]
\centering
\includegraphics[width=4in]{fig4}
\caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr gyrochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}. The bimodal rotation period distribution is clear, with peaks at $\Delta \log P_{rot}$ of -0.19 and 0.21 days. A two-Gaussian model was fit to this distribution (blue line).
Approximately 50\% more stars are present in the slower, nominally older peak (right).}
\label{fig:diff}
\end{figure}

%%%%%
\begin{figure}[]
\centering
\includegraphics[width=4in]{fig5}
\caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green). Median values for the two distributions are shown (vertical dashed lines), which yield a marginal 2.8$\sigma$ difference.
}
\label{fig:pm}
\end{figure}

\clearpage

The difference (in log period) between the observed rotation and the 600 Myr gyrochrone for the 440 likely main sequence stars is shown in Figure \ref{fig:diff}. Despite combining stars of all temperatures, the bimodality is clearly seen in this log period space. A two Gaussian model was fit to this data, which found peaks in the two distributions of $-0.19 \pm 0.01$ and $0.21 \pm 0.01$ days.  262 stars had periods longer than the gyrochrone model (right peak) and 178 slower than the model (left peak). This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample of stars with non-zero proper motions.

The favored explanation for the bimodal period distribution by \citet{mcquillan2013} was an age effect, with nearby stars having a bimodal star formation history. This explanation was bolstered by their observation that stars in the two period groups have differing distributions of proper motions, indicating they belonged to kinematically separate groups. This measurement is replicated in Figure \ref{fig:pm}, which shows the total proper motion distribution for stars above and below the 600 Myr gyrochrone. Stars above the gyrochrone (slower rotators, nominally older) have a median total proper motion of 15.4 mas/yr, while those below (faster rotators, younger)  have a median of 11.3 mas/yr. This difference in kinematics versus rotation period is in the same direction observed by \citet{mcquillan2013}. As the typical error in the  Gaia DR1 total proper motion is $\sim$1.4 mas/yr for this sample, this is a statistically marginal (2.8$\sigma$) difference.  Similarly, the Kolmogorov-Smirnov statistic for these two samples is 0.14, which does not rule out the null hypothesis that they are drawn from the same distribution. The slower rotating sample (black in Figure \ref{fig:pm}) appears to have a bimodal distribution in proper motion, indicating possibly significant contamination from a younger, lower proper motion population that is consistent with the rapidly rotating stars. Further investigation of rotation and kinematics of nearby stars using larger samples from future Gaia data releases is needed.

@@

Finally, this exploratory work has highlighted the utility of using astrometric data from Gaia combined with detailed light curve statistics from \Kepler to reveal hidden substructure in the properties of field stars. Looking forward to the astrometric precision of future Gaia data releases, this combination will be effective at separating dwarf stars from subgiants for nearly the entire \Kepler and K2 databases, and enable accurate age maps for field stars.

%%%%%%%%%%%%%%%%%
\acknowledgments
Special thanks to Kevin Covey, Jennifer van Saders, Sean Matt, Travis Metcalfe, and Aleks Scholz for their helpful discussions that motivated publishing this work. JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.

@@

\begin{thebibliography}{}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

@@

\bibitem[{{Barnes} \& {Kim}(2010)}]{barnes_kim2010}
{Barnes}, S.~A., \& {Kim}, Y.-C. 2010, \apj, 721, 675

@@

>>>>

\documentclass[manuscript, letterpaper]{aastex6}
%\documentclass[twocolumn, trackchanges]{aastex6}
%\documentclass[twocolumn]{aastex6}

@@

\altaffiltext{1}{Department of Physics \& Astronomy, Western Washington University, 516 High St., Bellingham, WA 98225, USA}
\altaffiltext{2}{NSF Astronomy and Astrophysics Postdoctoral Fellow}

@@

%tgas = J/ApJS/211/24/table1 \& I/337/tgas
%gaia = J/ApJS/211/24/table1 \& I/337/gaia
\begin{figure*}[]
\centering
\includegraphics[width=6in]{fig1}
\caption{
Rotation period distribution for 33,855 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots). The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$25 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}. The subsample of 1,299 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample. 
}
\label{fig:all}
\end{figure*}

@@

@@

%%%%%
\begin{figure}[]
\centering
\includegraphics[width=3.5in]{fig2}
\caption{Hertzsprung--Russell (HR) diagram using temperatures from \citet{mcquillan2014} and Gaia DR1 $G$-band absolute magnitudes for the 894 stars that pass photometric and parallax quality cuts described in the text. Points are colored by their measured \Kepler rotation periods. Two isochrones from \citet{bressan2012} are shown, with ages of 300 Myr and 4 Gyr (solid and dashed lines). Main sequence stars are selected as those between the 300 Myr isochrone and the isochrone shifted up by 1 magnitude (red line).
}
\label{fig:HR}
\end{figure}

@@

%%%%%
\begin{figure*}[]
\centering
\includegraphics[width=6in]{fig3}
\caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles). 
The full \Kepler--Gaia matched sample is shown for reference (small black dots). A bimodality in rotation periods initially discovered for M dwarfs by \citet{mcquillan2013}, extends the full range of temperatures in the \Kepler--Gaia main sequence sample shown here.
A \citet{meibom2011} 600 Myr gyrochonology-isochrone (gyrochrone) traces the bimodality midpoint up to 6000 K (red solid line), but deviates from the isochrone sharply to $\sim$6200 K (red dotted line). A log-linear extrapolation of the isochrone at 6000K to 6500 K (red dashed line) continues to track the bimodality to hotter temperatures, and roughly follows a line of constant Rossby number. 
}
\label{fig:gyro}
\end{figure*}

The minimum in the bimodal period distribution in Figure \ref{fig:gyro} can be traced using a gyrochonology isochrone (colloquially known as a ``gyrochrone'').  Many different studies have produced competing gyrochrone models, each with unique morphologies at the hot and cool star regimes \citep[e.g.][]{barnes2007,mamajek2008,meibom2011,angus2015}. A 600 Myr gyrochrone from \citet{meibom2011}, converted from $B-V$ color to temperature using the the transformation from \citet{sekiguchi2000}, was determined by eye to approximately trace the period minima from 3500 K to 6000 K. As shown in Figure \ref{fig:gyro}, this model (as with most gyrochronology models) turns down in period sharply for stars hotter than $\sim$6000 K. A log-linear extrapolation of the 600 Myr gyrochrone at 6000 K continues to trace the period bimodality up to 6500 K, and roughly traces a line of constant Rossby number assuming a local convective turnover timescale such as from \citet{barneskim2010}.

%%%%%
\begin{figure}[]
\centering
\includegraphics[width=3.5in]{fig4}
\caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr gyrochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro} \added{for the 440 main sequence selected stars (thick black line) and the initial sample of 894 stars with good TGAS detections (thin green line).}  The bimodal rotation period distribution \added{for main sequence stars} is clear, with peaks at $\Delta \log P_{rot}$ of -0.19 and 0.21 \replaced{days}{dex} from a two-Gaussian model fit (blue line). 
Approximately 50\% more stars are present in the slower, nominally older peak (right).
}
\label{fig:diff}
\end{figure}

The difference (in log period) between the observed rotation and the 600 Myr gyrochrone \deleted{for the 440 likely main sequence stars} is shown in Figure \ref{fig:diff}. Despite combining stars of all temperatures, the bimodality is clearly seen in this log period space \added{for the 440 likely main sequence stars}. A two Gaussian model was fit to this data, which found peaks in the two distributions of $-0.19 \pm 0.01$ and $0.21 \pm 0.01$ \replaced{days}{dex}.  262 stars had periods longer than the gyrochrone model (right peak) and 178 slower than the model (left peak). This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample of stars with non-zero proper motions. \added{The whole sample of 894 stars with good TGAS detections does not show this bimodality in Figure \ref{fig:diff}, demonstrating the importance of culling subgiants from the sample.}

%%%%%
\begin{figure}[]
\centering
\includegraphics[width=3.5in]{fig5a}
\includegraphics[width=3.5in]{fig5b}
\caption{Top: Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black), and below (rotating faster, younger stars) the model (green). \deleted{Median values for the two distributions are shown (vertical dashed lines), which yield a marginal 2.8$\sigma$ difference.}
\added{Bottom: Total proper motion as a function of the residual log rotation period shown in Figure \ref{fig:diff}, with the same color scheme for faster and slower rotating stars as the top panel.}
}
\label{fig:pm}
\end{figure}

The favored explanation for the bimodal period distribution by \citet{mcquillan2013} was an age effect, with nearby stars having a bimodal star formation history. This explanation was bolstered by their observation that stars in the two period groups have differing distributions of proper motions, indicating they belonged to kinematically separate groups. This measurement is replicated in Figure \ref{fig:pm}, which shows the total proper motion distribution for stars above and below the 600 Myr gyrochrone. Stars above the gyrochrone (slower rotators, nominally older) have a median total proper motion of 15.4 mas/yr, while those below (faster rotators, younger)  have a median of 11.3 mas/yr. This difference in kinematics versus rotation period is in the same direction observed by \citet{mcquillan2013}. \deleted{As the typical error in the  Gaia DR1 total proper motion is $\sim$1.4 mas/yr for this sample, this is a statistically marginal (2.8$\sigma$) difference.  Similarly, }The Kolmogorov-Smirnov statistic for these two samples is 0.14, which does not rule out the null hypothesis that they are drawn from the same distribution. The slower rotating sample (black in Figure \ref{fig:pm}) appears to have a bimodal distribution in proper motion, indicating possibly significant contamination from a younger, lower proper motion population that is consistent with the rapidly rotating stars. \added{Figure \ref{fig:pm} also shows the proper motion as a function of the residual rotation period defined in Figure \ref{fig:diff}. Despite the small sample size, this distribution highlights the complex relationship between ages derived from gyrochronology and galactic kinematics, including bimodal structure for the proper motion of slower rotating stars.} Further investigation of rotation and kinematics of nearby stars using larger samples from future Gaia data releases is needed.

@@

Finally, this exploratory work has highlighted the utility of using astrometric data from Gaia combined with detailed light curve statistics from \Kepler to reveal hidden substructure in the properties of field stars. Looking forward to the astrometric precision of future Gaia data releases, this combination will be effective at separating dwarf stars from subgiants for nearly the entire \Kepler and K2 databases, and enable accurate age maps for field stars.
% super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}.

%%%%%%%%%%%%%%%%%
\acknowledgments
Special thanks to Kevin Covey, Jennifer van Saders, Sean Matt, Travis Metcalfe, and Aleks Scholz for their helpful discussions that motivated publishing this work.
\added{Thanks to the anonymous referee for their helpful comments that improved this manuscript. Jay Strader provided a helpful reference for the correct notation for transcendental functions, \citet{matta2011}.}
JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.

@@

%\bibliography{/Users/davenpj3/Dropbox/references.bib}
\begin{thebibliography}{}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

@@

\bibitem[{{Barnes} \& {Kim}(2010)}]{barneskim2010}
{Barnes}, S.~A., \& {Kim}, Y.-C. 2010, \apj, 721, 675

@@

\bibitem[{{Matta} {et~al.}(2011){Matta}, {Massa}, {Gubskaya}, \&
  {Knoll}}]{matta2011}
{Matta}, C.~F., {Massa}, L., {Gubskaya}, A.~V., \& {Knoll}, E. 2011, Journal of
  Chemical Education, 88, 67
